---
title: "Compare Gene Loadings from AES-PCA for zebrafish and mouse"
author: "Nhi Hin"
date: "2020-08-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r Setup, include=FALSE}
# Load packages:
library(here)
library(magrittr)
library(dplyr)
library(biomaRt)
library(AnnotationHub)
library(purrr)
library(tidyr)
library(org.Mm.eg.db)
library(org.Dr.eg.db)
```


## Introduction

- Previously, AES-PCA was run using the Hallmark gene set collection on 
the fAD-like zebrafish and 5XFAD mouse datasets. 

- The results of AES-PCA include sparse PC values (PC1 and PC2) in addition to 
loadings that indicate the genes contributing most to the variation in 
gene set activity across samples. 

- Here, we will compare the gene loadings from two datasets to see whether 
gene set activity tends to be driven by the same or by different genes. 

- This is important to know as it would highlight the utility of looking 
at changes at the gene set level (more broader state compared to 
gene level), while also showing how AES-PCA allows us to go into the 
gene level as well. 

## 1. Import Data

- The AES-PCA loadings for PC1 were saved in the `pathwayPCA-mouse.Rmd` and 
`pathwayPCA-zebrafish.Rmd` documents. Here we will load in the saved objects.

```{r}
loadings_zeb <- readRDS(here("data", "datasets", "zebrafish", 
                              "geneLoadings.rds"))

loadings_mouse <-  readRDS(here("data", "datasets", "mouse", "GSE140286", 
                              "geneLoadings.rds"))
```

- Confirm that the same gene sets are in the datasets.

```{r}
names(loadings_zeb$q96VsWT_6mth) %in% names(loadings_mouse$XFADVsWT_3mth)
```

- There is one `FALSE` indicating a mismatch (spelling mistake) from using 
different versions of MSigDB Hallmark gene sets for zebrafish and mouse. 

```{r}
names(loadings_zeb$q96VsWT_6mth)[!names(loadings_zeb$q96VsWT_6mth) %in% names(loadings_mouse$XFADVsWT_3mth)]

loadings_mouse$XFADVsWT_3mth$HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY
```

```{r}
names(loadings_zeb$q96VsWT_6mth) %<>%
  gsub(x = ., pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
       replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")

names(loadings_zeb$q96vsWT_24mth) %<>%
  gsub(x = ., pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
       replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")
```

## 2. Convert gene IDs

- We need to convert gene IDs from mouse to zebrafish and vice versa.

```{r}
# Mouse entrez and ensembl ids:
mEns <- org.Mm.egENSEMBL %>%
  as.data.frame %>%
  set_colnames(c("m_entrezgene","m_ensembl"))

# Create a data.frame to map between human entrezgenes and zebrafish ensembl IDs.
# BioMart only includes homolog mappings for ensembl IDs which is why we need to 
# retrieve human ensembl IDs, then join to the humanEntrezEns data.frame,
# in order to get the desired human entrezgenes to zebrafish ensembl ID mapping. 
mMart <- useMart("ENSEMBL_MART_ENSEMBL", "mmusculus_gene_ensembl")
getFromBiomart <- c("ensembl_gene_id", "drerio_homolog_ensembl_gene")
mAndZebEnsGenes <- getBM(getFromBiomart, values = unique(mEns$m_ensembl), 
                           mart = mMart) %>%
  set_colnames(c("mouse_ensembl", "zeb_ensembl")) %>%
  dplyr::filter(complete.cases(.),
                zeb_ensembl != "")
```

- Try the other way around too in case there are some zebrafish-mouse 
conversions missed by going from mouse to zebrafish gene IDs.

```{r}
# Zebrafish entrez and ensembl ids:
zEns <- org.Dr.egENSEMBL %>%
  as.data.frame %>%
  set_colnames(c("z_entrezgene","z_ensembl"))

# Create a data.frame to map between human entrezgenes and zebrafish ensembl IDs.
# BioMart only includes homolog mappings for ensembl IDs which is why we need to 
# retrieve human ensembl IDs, then join to the humanEntrezEns data.frame,
# in order to get the desired human entrezgenes to zebrafish ensembl ID mapping. 
zmart <- useMart("ENSEMBL_MART_ENSEMBL", "drerio_gene_ensembl")
getFromBiomart <- c("ensembl_gene_id", "mmusculus_homolog_ensembl_gene")
zAndMEnsGenes <- getBM(getFromBiomart, values = unique(zEns$z_ensembl), 
                           mart = zmart) %>%
  set_colnames(c("zeb_ensembl", "mouse_ensembl")) %>%
  dplyr::filter(complete.cases(.),
                mouse_ensembl != "")
```

```{r}
dim(mAndZebEnsGenes)
dim(zAndMEnsGenes)

table(mAndZebEnsGenes$zeb_ensembl %in% zAndMEnsGenes$zeb_ensembl)
```

```{r}

loadings_zeb %<>% lapply(function(x){
  x %>% lapply(function(y){
    y %>% left_join(zAndMEnsGenes, by = c("gene_id"="zeb_ensembl"))
  })
})

# loadings_zeb$q96vsWT_24mth %<>% lapply(function(x){
#   x %>%left_join(zAndMEnsGenes, by = c("gene_id"="zeb_ensembl"))
# })


loadings_mouse %<>% lapply(function(x){
  x %>% lapply(function(y){
    y %>% left_join(mAndZebEnsGenes, by = c("gene_id"="mouse_ensembl"))
  })
})
```



## 2. Comparisons

### 2.1. Common genes driving changes in young adult fAD-like and 5XFAD brains

```{r}
# Loadings of the zebrafish, using list-column workflow on purrr
youngLoadings_zeb <- bind_rows(loadings_zeb$q96VsWT_6mth, .id = "Geneset") %>%
  dplyr::group_by(Geneset) %>%
  tidyr::nest() %>%
  dplyr::rename(data_zeb = data)

# Loadings of mouse
youngLoadings_mouse <- bind_rows(loadings_mouse$XFADVsWT_3mth, 
                                 .id="Geneset") %>%
  dplyr::group_by(Geneset) %>%
  tidyr::nest() %>%
  dplyr::rename(data_mouse = data)

# Tidy data.frame
youngLoadings_all <- full_join(youngLoadings_zeb, 
                               youngLoadings_mouse, by = "Geneset") %>%
  mutate(overlapping_genes_mouse = map2(data_zeb, data_mouse, function(x,y){
    inner_join(x, y, by = c("mouse_ensembl"="gene_id"))
  })) 
youngLoadings_all$overlapping_genes_mouse %<>% set_names(
  youngLoadings_all$Geneset
)

# Get the genes which contribute to PC1 in both datasets.
conservedGenes <- youngLoadings_all$overlapping_genes_mouse %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::filter(gene_id == zeb_ensembl) %>% 
  dplyr::select(Geneset, gene_id, PC1.x, 
                mouse_ensembl, PC1.y,
                gene_name.y, description.y) %>%
  dplyr::rename(gene_name = gene_name.y,
                description = description.y,
                zeb_ensembl = gene_id,
                PC1_zeb = PC1.x,
                PC1_mouse = PC1.y) 
```

### 2.2. Common genes driving changes in aged brains

- Same thing as above but for the older age group.

```{r}
# Loadings of the zebrafish, using list-column workflow on purrr
agedLoadings_zeb <- bind_rows(loadings_zeb$q96vsWT_24mth, .id = "Geneset") %>%
  dplyr::group_by(Geneset) %>%
  tidyr::nest() %>%
  dplyr::rename(data_zeb = data)

# Loadings of mouse
agedLoadings_mouse <- bind_rows(loadings_mouse$XFADvsWT_12mth, 
                                 .id="Geneset") %>%
  dplyr::group_by(Geneset) %>%
  tidyr::nest() %>%
  dplyr::rename(data_mouse = data)

# Tidy data.frame
agedLoadings_all <- full_join(agedLoadings_zeb, 
                               agedLoadings_mouse, by = "Geneset") %>%
  mutate(overlapping_genes_mouse = map2(data_zeb, data_mouse, function(x,y){
    inner_join(x, y, by = c("mouse_ensembl"="gene_id"))
  })) 
agedLoadings_all$overlapping_genes_mouse %<>% set_names(
  agedLoadings_all$Geneset
)

# Get the genes which contribute to PC1 in both datasets.
conservedGenes_aged <- agedLoadings_all$overlapping_genes_mouse %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::filter(gene_id == zeb_ensembl) %>% 
  dplyr::select(Geneset, gene_id, PC1.x, 
                mouse_ensembl, PC1.y,
                gene_name.y, description.y) %>%
  dplyr::rename(gene_name = gene_name.y,
                description = description.y,
                zeb_ensembl = gene_id,
                PC1_zeb = PC1.x,
                PC1_mouse = PC1.y) 
```

## 3. Summary of Results

- Overall, it seems that the genes driving variation between samples 
in the fAD-like zebrafish and 5XFAD mouse datasets are mostly different, 
with only a few similarities. 

- The similar genes at young age are shown below.

```{r}
conservedGenes
```

- Old age:

```{r}
conservedGenes_aged
```

- Limitations are that the young age groups may not be directly comparable 
or the most equivalent due to different life spans of zebrafish and 
mice, so this is an approximation. However, the results overall 
do seem to indicate that the pathological states / changes occurring 
due to the fAD-like and 5XFAD mutations differ quite significantly, despite 
some of the gene sets being in common. 

- The significance of this is that we are able to detect some similarities 
between these two animal models in terms of the types of biological 
activities that are altered as a result of the respective mutations. 
However, the genes driving these changes largely differ. 

## Export Objects

- Lists of genes which contribute to AES PC1 loadings similarly in 
fAD-like zebrafish and 5XFAD mouse brains. 

```{r eval=FALSE}
conservedGenes %>% saveRDS(here("data", "loadings", 
                                "conservedGenes_young.rds"))

conservedGenes_aged %>% saveRDS(here("data", "loadings", 
                                     "conservedGenes_aged.rds"))
```

