---
title: "Comparison of 5xFAD mouse and fAD-like zebrafish brain transcriptomes"
author: "Nhi Hin"
date: "2020-05-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
library(here)
library(dplyr)
library(ggplot2)
library(magrittr)
theme_set(theme_bw())
```


## 0. Introduction/Aims

- Perform a preliminary comparison of the 5xFAD mouse and 
fAD-like zebrafish datasets through comparing their gene set testing results. 
This is a straightforward approach that should tell us how comparable 
they are at first glance. 

- Establish if there are particular ages of the mice that 
are more comparable to zebrafish. We have various ages of mouse brains 
ranging from 1 month old to 12 months old. (zebrafish 6 and 24 months old). 

## 1. Import gene set test results

- Previously gene set testing was done using the MSigDB collection Hallmark 
gene set collection. 

- The `combinedGSEA` function includes the methods *fry*, *camera* and 
*fgsea* and combines the raw p-values using Wilkinson's method before 
adjusting for multiple tests. 

- The results are imported below.

```{r}
# 6 and 24 months old, WT and Q96_K97del/+ mutant
zebResults <- readRDS(here("data", "datasets", "zebrafish", 
                             "gsea.rds"))

# 1, 2, and 4 months old, 5XFAD only
mouResults1 <- readRDS(here("data", "datasets", "mouse", 
                            "GSE97113", "gsea.rds"))

# 3, 6, and 12 months old, WT and 5XFAD
mouResults2 <- readRDS(here("data", "datasets", "mouse", 
                            "GSE140286", "gsea.rds"))

# 6 and 11 months old, WT and 5XFAD
mouResults3 <- readRDS(here("data", "datasets", "mouse", 
                            "GSE142633", "gsea.rds"))
```

## 2. Proportions of up and down-regulated genes in Hallmark gene sets

- The `roast` method for gene set testing conveniently gives us the proportion 
of genes which are increased (t > 2) or decreased (t < -2) in expression 
for each comparison. 

- Note that the `mroast` results were not included when combining p-values 
with Wilkinson's method (only `fry`, `fgsea` and `camera` were included). 

- The rankings for gene sets using the methods `mroast` and `fry` are generally 
very similar, but `fry` p-values are smaller due to the infinite rotations 
approximation. 

```{r}
# Extract proportions of up and down genes from mroast for zebrafish
zeb_facet1 <- zebResults$indivTest$normoxia_6mth_mutant_vs_wt$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "zeb_effectOfMutation_6mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown) %>%
  mutate(Geneset = gsub(x = Geneset,
                        pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY",
                        replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY"))
zeb_facet1$fdr <- zebResults$combTest$normoxia_6mth_mutant_vs_wt$fdr
zeb_facet1 %<>% mutate(significant = fdr < 0.05)

zeb_facet2 <- zebResults$indivTest$normoxia_24mth_mutant_vs_wt$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "zeb_effectOfMutation_24mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)%>%
  mutate(Geneset = gsub(x = Geneset,
                      pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY",
                      replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY"))
zeb_facet2$fdr <- zebResults$combTest$normoxia_24mth_mutant_vs_wt$fdr
zeb_facet2 %<>% mutate(significant = fdr < 0.05)

# Dataset 2 (different ages)
mou2_facet1 <- mouResults2$indivTest$effectOf5XFAD_3mth$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "mou_effectOfMutation_3mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)
mou2_facet1$fdr <- mouResults2$combTest$effectOf5XFAD_3mth$fdr
mou2_facet1 %<>% mutate(significant = fdr<0.05)

mou2_facet2 <- mouResults2$indivTest$effectOf5XFAD_6mth$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "mou_effectOfMutation_6mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)
mou2_facet2$fdr <- mouResults2$combTest$effectOf5XFAD_6mth$fdr
mou2_facet2 %<>% mutate(significant = fdr<0.05)

mou2_facet3 <- mouResults2$indivTest$effectOf5XFAD_12mth$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "mou_effectOfMutation_12mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)
mou2_facet3$fdr <- mouResults2$combTest$effectOf5XFAD_12mth$fdr
mou2_facet3 %<>% mutate(significant = fdr<0.05)

# Dataset 3 - 6 and 11 months old. 
mou3_facet1 <- mouResults3$indivTest$effectOf5XFAD_6mth$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "mou_effectOfMutation_6mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)
mou3_facet1$fdr <- mouResults3$combTest$effectOf5XFAD_6mth$fdr
mou3_facet1 %<>% mutate(significant = fdr<0.05)

mou3_facet2 <- mouResults3$indivTest$effectOf5XFAD_11mth$mroast %>%
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>%
  dplyr::mutate(contrast = "mou_effectOfMutation_11mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down) %>%
  dplyr::select(-PropUp, -PropDown)
mou3_facet2$fdr <- mouResults3$combTest$effectOf5XFAD_11mth$fdr
mou3_facet2 %<>% mutate(significant = fdr<0.05)

```

- Plot both zebrafish age groups as well as mice from the dataset with 
ages 3, 6 and 12 months. 

```{r}
butterfly <- bind_rows(zeb_facet1, zeb_facet2, 
                       mou2_facet1, mou2_facet2,mou2_facet3) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " "),
                Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK", 
                               replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset",
                           "NGenes",
                           "contrast",
                           "significant")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, 
                                   levels = c("notSig", "down", "up")),
                contrast = factor(contrast, 
                                  levels = c("zeb_effectOfMutation_6mth",
                                             "zeb_effectOfMutation_24mth",
                                             "mou_effectOfMutation_3mth", 
                                             "mou_effectOfMutation_6mth", 
                                             "mou_effectOfMutation_12mth")))%>%
  dplyr::filter(!is.na(direction)) %>%
  dplyr::mutate(significant = case_when(
    direction == "up" ~ " ",
    direction == "down" ~ " ",
    direction == "notSig" & significant == TRUE ~ "*",
    direction == "notSig" & significant == FALSE ~ " ")) %>% 
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  facet_grid(~contrast, scales = "free") +
  #ggpol::facet_share(~contrast, scales = "free", strip.position = "left") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip() +
    geom_text(aes(label=significant),vjust=0.77, hjust = 0)+
  labs(x = "Number of genes", y = "Hallmark gene set",
       fill = "Direction")

butterfly
```

- Plot the young zebrafish and young mice only.

```{r}
butterfly_young <- bind_rows(zeb_facet1, mou2_facet1) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " "),
                Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK", 
                               replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset",
                           "NGenes",
                           "contrast",
                           "significant")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, 
                                   levels = c("notSig", "down", "up")),
                contrast = factor(contrast, 
                                  levels = c("zeb_effectOfMutation_6mth",
                                             "mou_effectOfMutation_3mth")))%>%
  dplyr::filter(!is.na(direction)) %>%
  dplyr::mutate(significant = case_when(
    direction == "up" ~ " ",
    direction == "down" ~ " ",
    direction == "notSig" & significant == TRUE ~ "*",
    direction == "notSig" & significant == FALSE ~ " ")) %>% 
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  facet_grid(~contrast, scales = "free") +
  #ggpol::facet_share(~contrast, scales = "free", strip.position = "left") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip() +
    geom_text(aes(label=significant),vjust=0.77, hjust = 0)+
  labs(x = "Number of genes", y = "Hallmark gene set",
       fill = "Direction")

butterfly_young

#export::graph2ppt(butterfly_young, here("data", "initial_comparisons",
#                                        "fig", "butterfly_young"))

```

- Plot the aged zebrafish and aged mice from both datasets (11 and 12 
months). 

```{r}
butterfly_aged <- bind_rows(zeb_facet2, mou3_facet2, mou2_facet3 ) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " "),
                Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK", 
                               replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset",
                           "NGenes",
                           "contrast",
                           "significant")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, 
                                   levels = c("notSig", "down", "up")),
                contrast = factor(contrast, 
                                  levels = c("zeb_effectOfMutation_24mth",
                                             "mou_effectOfMutation_11mth",
                                             "mou_effectOfMutation_12mth")))%>%
  dplyr::filter(!is.na(direction)) %>%
  dplyr::mutate(significant = case_when(
    direction == "up" ~ " ",
    direction == "down" ~ " ",
    direction == "notSig" & significant == TRUE ~ "*",
    direction == "notSig" & significant == FALSE ~ " ")) %>% 
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  facet_grid(~contrast, scales = "free") +
  #ggpol::facet_share(~contrast, scales = "free", strip.position = "left") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip() +
    geom_text(aes(y = NGenes, label=significant),vjust=0.77, 
              hjust = 0, size = 5.5)+
  labs(y = "Number of genes", x= "Hallmark gene set",
       fill = "Direction")+
  ylim(0,260)

butterfly_aged

# export::graph2ppt(butterfly_aged, here("data", "initial_comparisons",
#                                        "fig", "butterfly_aged"))

```






```{r}
butterfly2 <- bind_rows(zeb_facet1, zeb_facet2, mou2_facet1, mou2_facet2,mou2_facet3) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, pattern = "_", replacement = " "),
                Geneset = gsub(x = Geneset, pattern = "HALLMARK", 
                               replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset","NGenes","contrast", "significant")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, 
                                   levels = c("notSig", "down", "up")),
                contrast = factor(contrast, 
                                  levels = c("zeb_effectOfMutation_6mth",
                                             "zeb_effectOfMutation_24mth",
                                             "mou_effectOfMutation_3mth", 
                                             "mou_effectOfMutation_6mth", 
                                             "mou_effectOfMutation_12mth")))%>%
  dplyr::filter(!is.na(direction)) %>%
  dplyr::mutate(significant = case_when(
    direction == "up" ~ " ",
    direction == "down" ~ " ",
    direction == "notSig" & significant == TRUE ~ "*",
    direction == "notSig" & significant == FALSE ~ " ")) %>% 
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  facet_grid(~contrast, scales = "free") +
  #ggpol::facet_share(~contrast, scales = "free", strip.position = "left") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  geom_text(aes(label=significant))+
  coord_flip() +
  labs(x = "Number of genes", y = "Hallmark gene set",
       fill = "Direction")

butterfly2
```



```{r}
# extract up/downregulated gene proportions from mroast
x_props <- gseaResults_mouse_h$indivTest$effectOfMutation_3mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_3mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

x_props2 <- gseaResults_mouse_ire$indivTest$effectOfMutation_3mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_3mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

x_props <- bind_rows(x_props, x_props2)
x_fry_with_props <- x_props

# join to fry results
x_fry_with_props <- gseaResults_mouse_h$indivTest$effectOfMutation_3mth$fry %>%
  dplyr::filter(FDR.Mixed < 0.05) %>%
  dplyr::select(Geneset) %>%
  left_join(x_props)


# extract proportions from mroast
y_props <- gseaResults_mouse_h$indivTest$effectOfMutation_6mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_6mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

y_props2 <- gseaResults_mouse_ire$indivTest$effectOfMutation_6mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_6mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

y_props <- bind_rows(y_props, y_props2)

# join to fry results
# y_fry_with_props <- gseaResults_mouse_h$indivTest$effectOfMutation_6mth$fry %>%
#   #dplyr::filter(FDR.Mixed < 0.05) %>%
#   dplyr::select(Geneset) %>%
#   left_join(y_props)

y_fry_with_props <- y_props

####### slightly dodge but OK 
z_props <- gseaResults_mouse_h$indivTest$effectOfMutation_12mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_12mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

z_props2 <- gseaResults_mouse_ire$indivTest$effectOfMutation_12mth$mroast %>% 
  dplyr::select(Geneset, NGenes, PropUp, PropDown) %>% 
  dplyr::mutate(contrast="effectOfMutation_12mth",
                up = NGenes * PropUp,
                down = NGenes * PropDown,
                notSig = NGenes - up - down
  ) %>%
  dplyr::select(-PropUp, -PropDown)

z_props <- bind_rows(z_props, z_props2)
z_fry_with_props <- z_props

# join to fry results
# z_fry_with_props <- gseaResults_mouse_h$indivTest$effectOfMutation_12mth$fry %>%
#   #dplyr::filter(FDR.Mixed < 0.05) %>%
#   dplyr::select(Geneset) %>%
#   left_join(z_props)

z_fry_with_props



# create the plot
# adapted from https://stackoverflow.com/questions/4559229/drawing-pyramid-plot-using-r-and-ggplot2
butterfly <- bind_rows(x_fry_with_props, y_fry_with_props, z_fry_with_props) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, pattern = "_", replacement = " "),
                Geneset = gsub(x = Geneset, pattern = "HALLMARK", replacement = "")) %>%
  reshape2::melt(id.vars=c("Geneset","NGenes","contrast")) %>%
  dplyr::arrange(desc(NGenes)) %>% 
  dplyr::rename(direction = variable, ngenes = value) %>%
  dplyr::mutate(Geneset = factor(Geneset, levels = unique(Geneset)),
                direction = factor(direction, levels = c("notSig", "down", "up")),
                contrast = factor(contrast, levels = c("effectOfMutation_3mth","effectOfMutation_6mth", "effectOfMutation_12mth"))) %>%
  ggplot(aes(x = Geneset, y = ngenes, fill = direction)) +
  geom_col(width=0.6) +
  facet_wrap(~contrast, scales = "free") +
  #ggpol::facet_share(~contrast, scales = "free", strip.position = "left") +
  scale_fill_manual(values=c("#DDDDDD","cornflowerblue","#FE4365"))+
  coord_flip() +
  theme(axis.line.y = element_blank())



butterfly

export::graph2ppt(butterfly, here("R/Mouse/fig/butterfly_effectOfMutation"))
```


