---
title: "AES-PCA Loadings - Figures"
author: "Nhi Hin"
date: "2020-08-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
# Load packages:
library(here)
library(magrittr)
library(dplyr)
library(biomaRt)
library(AnnotationHub)
library(purrr)
library(tidyr)
library(org.Mm.eg.db)
library(org.Dr.eg.db)
library(ggplot2)
library(export)
```


## Introduction

- AES-PCA was run for the fAD-like and 5XFAD mouse model datasets, 
returning the sparse PC values (PC1 and PC2) as well as gene loadings 
for these.

- Here, we will further explore the differences in the genes contributing to the 
variation across PC1 (gene loadings).

## 1. Import Data

- The AES-PCA loadings for PC1 were saved in the `pathwayPCA-mouse.Rmd` and 
`pathwayPCA-zebrafish.Rmd` documents. Here we will load in the saved objects.

```{r}
loadings_zeb <- readRDS(here("data", "datasets", "zebrafish", 
                              "geneLoadings.rds"))

loadings_mouse <-  readRDS(here("data", "datasets", "mouse", "GSE140286", 
                              "geneLoadings.rds"))
```

- Confirm that the same gene sets are in the datasets.

```{r}
names(loadings_zeb$q96VsWT_6mth) %in% names(loadings_mouse$XFADVsWT_3mth)
```

- There is one `FALSE` indicating a mismatch (spelling mistake) from using 
different versions of MSigDB Hallmark gene sets for zebrafish and mouse. 

```{r}
names(loadings_zeb$q96VsWT_6mth)[!names(loadings_zeb$q96VsWT_6mth) %in% names(loadings_mouse$XFADVsWT_3mth)]

loadings_mouse$XFADVsWT_3mth$HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY
```

```{r}
names(loadings_zeb$q96VsWT_6mth) %<>%
  gsub(x = ., pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
       replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")

names(loadings_zeb$q96vsWT_24mth) %<>%
  gsub(x = ., pattern = "HALLMARK_REACTIVE_OXIGEN_SPECIES_PATHWAY", 
       replacement = "HALLMARK_REACTIVE_OXYGEN_SPECIES_PATHWAY")
```

- Import in the Gene Set enrichment analysis results indicating the 
gene sets found to be significant in each comparison of each
dataset.

```{r}
zebResults <- readRDS(here("data", "datasets", "zebrafish", 
                             "gsea.rds"))

mouseResults <- readRDS(here("data", "datasets", "mouse", 
                            "GSE140286", "gsea.rds"))
```


## 2. Zebrafish Dataset

- The significantly enriched gene sets (determined from previous gene 
set analysis) will have their AES-PC1 values plotted below. 

- Note: As AES-PCA creates sparse PCAs, the loadings of many of the genes will 
have been set to 0. 

- **Young zebrafish (6 months old)**

```{r}
# Gene sets which were significantly enriched in the young 
# fAD-like mutant vs wt comparison
sigInYoungZeb <- zebResults$combTest$normoxia_6mth_mutant_vs_wt %>%
  dplyr::filter(fdr < 0.05)

loadingsPlot_youngZeb <- loadings_zeb$q96VsWT_6mth %>% bind_rows(.id = "Geneset") %>% 
  dplyr::distinct(Geneset, gene_id, PC1, 
                  gene_name, .keep_all = TRUE) %>% 
  dplyr::filter(Geneset %in% sigInYoungZeb$Geneset)%>%
  dplyr::arrange(Geneset, PC1) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " ")) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK",
                               replacement = "")) %>%
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)), y = PC1)) +
  geom_bar(stat="identity") +
  facet_wrap(~Geneset, scales = "free_y", ncol= 3) +
  coord_flip()+
  theme(aspect.ratio = 1) +
  labs(y = "PC1", x = "Gene loadings")

loadingsPlot_youngZeb
```

- **Aged zebrafish (24 months old)**

```{r}
sigInAgedZeb <- zebResults$combTest$normoxia_24mth_mutant_vs_wt %>%
  dplyr::filter(fdr < 0.05)

loadingsPlot_agedZeb <- loadings_zeb$q96vsWT_24mth %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::distinct(Geneset, gene_id, PC1, 
                  gene_name, .keep_all = TRUE) %>% 
  dplyr::filter(Geneset %in% sigInAgedZeb$Geneset)%>%
  dplyr::arrange(Geneset, PC1) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " ")) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK",
                               replacement = "")) %>%
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)), y = PC1)) +
  geom_bar(stat="identity") +
  facet_wrap(~Geneset, scales = "free_y", ncol= 3) +
  coord_flip()+
  theme(aspect.ratio = 1) +
  labs(y = "PC1", x = "Gene loadings")

loadingsPlot_agedZeb
```

## 3. Mouse Dataset

- **Young mice (3 months old)**

```{r}
sigInYoungMice <- mouseResults$combTest$effectOf5XFAD_3mth %>%
  dplyr::filter(fdr < 0.05) 

loadingsPlot_youngMice <- loadings_mouse$XFADVsWT_3mth %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::distinct(Geneset, gene_id, PC1, 
                  gene_name, .keep_all = TRUE) %>% 
  dplyr::filter(Geneset %in% sigInYoungMice$Geneset)%>%
  dplyr::arrange(Geneset, PC1) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " ")) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK",
                               replacement = "")) %>%
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)), y = PC1)) +
  geom_bar(stat="identity") +
  facet_wrap(~Geneset, scales = "free_y", nrow=3) +
  coord_flip()+
  #theme(aspect.ratio = 1) +
  labs(y = "PC1", x = "Gene loadings")

loadingsPlot_youngMice
```

- **Aged mice (12 months old)**

```{r}
sigInAgedMice <- mouseResults$combTest$effectOf5XFAD_12mth %>%
  dplyr::filter(fdr < 0.05) 

loadingsPlot_agedMice <- loadings_mouse$XFADvsWT_12mth %>% 
  bind_rows(.id = "Geneset") %>% 
  dplyr::distinct(Geneset, gene_id, PC1, 
                  gene_name, .keep_all = TRUE) %>% 
  dplyr::filter(Geneset %in% sigInAgedMice$Geneset)%>%
  dplyr::arrange(Geneset, PC1) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " ")) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK",
                               replacement = "")) %>%
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)), y = PC1)) +
  geom_bar(stat="identity") +
  facet_wrap(~Geneset, scales = "free_y", ncol= 3) +
  coord_flip()+
  #theme(aspect.ratio = 1) +
  labs(y = "PC1", x = "Gene loadings")

loadingsPlot_agedMice
```

## 4. Gene sets enriched in both datasets

- Import in the conserved genes to see if these overlap with the common 
gene sets. 

```{r}
conservedGenes <-  readRDS(here("data", "loadings", 
                                "conservedGenes_young.rds"))
```


```{r}
sigInYoungAll <- sigInYoungZeb %>%
  full_join(sigInYoungMice, by = c("Geneset"))%>%
  dplyr::filter(!is.na(id.x) , !is.na(id.y)) %>%
  .$Geneset

loadingsPlot_youngCommon <- list(Zebrafish = loadings_zeb$q96VsWT_6mth, 
                              Mouse = loadings_mouse$XFADVsWT_3mth) %>%
  lapply(function(x){
    x %>%
      bind_rows(.id = "Geneset") %>%
      dplyr::distinct(Geneset, gene_id, PC1, gene_name, .keep_all = TRUE) %>%
      dplyr::filter(Geneset %in% sigInYoungAll) %>%
      dplyr::arrange(Geneset)
  }) %>%
  bind_rows(.id = "Dataset") %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "_",
                               replacement = " ")) %>%
  dplyr::mutate(Geneset = gsub(x = Geneset, 
                               pattern = "HALLMARK",
                               replacement = "")) 
loadingsPlot_youngCommon

theme_set(theme_bw(base_size = 16))

geneset1 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" EPITHELIAL MESENCHYMAL TRANSITION")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single", width=0.3)) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=1)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset1 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset1.pdf"))
geneset1 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset1"))

###
geneset2 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" FATTY ACID METABOLISM")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=0.5)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset2 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset2.pdf"))
geneset2 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset2"))

###
geneset3 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" HEME METABOLISM")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=0.5)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset3 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset3.pdf"))
geneset3 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset3"))

###
geneset4 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" PI3K AKT MTOR SIGNALING")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=0.5)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset4 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset4.pdf"))
geneset4 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset4"))

###
geneset5 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" PROTEIN SECRETION")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=0.5)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset5 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset5.pdf"))
geneset5 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset5"))

###
geneset6 <- loadingsPlot_youngCommon %>%
  dplyr::filter(Geneset %in% c(" UV RESPONSE DN")) %>% 
  ggplot(aes(x = factor(gene_name, levels = unique(gene_name)),
             y = PC1)) +
  geom_col(position = position_dodge2(preserve = "single")) +
  facet_grid(~Geneset + Dataset, scales = "free_x") +
  coord_flip() +
  theme(aspect.ratio=0.5)+
  labs(y = "PC1", x = "Gene loadings") 
  
geneset6 %>% export::graph2pdf(here("data", "loadings", "fig", "geneset6.pdf"))
geneset6 %>% export::graph2ppt(here("data", "loadings", "fig", "geneset6"))


```


## Export Objects

```{r eval=FALSE}
loadingsPlot_youngZeb %>% export::graph2pdf(here("data", "loadings", 
                                                 "fig", 
                                                 "loadingsPlot_youngZeb.pdf"),
                                            width = 8, height = 11)

loadingsPlot_agedZeb %>% export::graph2pdf(here("data", "loadings", 
                                                 "fig", 
                                                 "loadingsPlot_agedZeb.pdf"),
                                            width = 8, height = 23)

loadingsPlot_youngMice %>% export::graph2pdf(here("data", "loadings", 
                                                 "fig", 
                                                 "loadingsPlot_youngMice.pdf"),
                                            width = 8, height = 11)

loadingsPlot_agedMice %>% export::graph2pdf(here("data", "loadings", 
                                                 "fig", 
                                                 "loadingsPlot_agedMice.pdf"),
                                            width = 8, height = 23)

loadingsPlot_youngCommon %>% export::graph2pdf(here("data", "loadings",
                                                    "fig", 
                                              "loadingsPlot_youngCommon.pdf"),
                                              width = 8, height = 11)

loadingsPlot_youngCommon %>% export::graph2ppt(here("data", "loadings",
                                                    "fig", 
                                              "loadingsPlot_youngCommon"),
                                              width = 8, height = 11)
```


