---
title: "PathwayPCA on 5XFAD Mouse Dataset"
author: "Nhi Hin"
date: "2020-06-01"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(pathwayPCA)
library(edgeR)
library(here)
library(ggplot2)
library(magrittr)
library(dplyr)
library(tibble)
```


### 1. Import gene sets

- According to the [PathwayPCA vignette](https://gabrielodom.github.io/pathwayPCA/articles/Supplement2-Importing_Data.html#overview) we need to set up a `pathwayCollection` object of the gene 
sets to be tested. Here I will use the Hallmark collection from MSigDB. 

- The `ens_h_mapped.rds` object was previously created from the Hallmark 
collection .gmt file from MSigDB (v.7.0). The entrezgene IDs were converted 
to Ensembl mouse IDs using homologous genes from BioMart. 

```{r}
hallmarkGenes <- readRDS(here("data", "genesets", "mouse",
                              "ens_h_mapped.rds"))
hallmarkNames <- names(hallmarkGenes)

hallmarkCol <- CreatePathwayCollection(
  sets_ls = hallmarkGenes,
  TERMS = hallmarkNames
)
```

### 2. Tidy sample/expression matrix

- The expression data must be in tidy format for inputting into 
*PathwayPCA*. 

```{r}
exprs <- readRDS(here("data", "datasets", "mouse", 
                      "GSE140286", "dge.rds")) %>%
  cpm(log=TRUE)

samples <- readRDS(here("data", "datasets", "mouse", 
                        "GSE140286", "dge.rds"))$samples
samples %<>% rownames_to_column("Sample")

tidyDat <- exprs %>% 
  as.data.frame %>%
  rownames_to_column("gene") %>% 
  TransposeAssay()

tidyDat[1:5, 1:20]  # preview
```

### 3. `Omics` data object

- The samples/expression tidy data and gene sets must now be placed inside an 
`Omics` object.

- We have to create different `Omics` objects to test different contrasts. 
*PathwayPCA* does not support complex designs like regular gene set testing 
methods utilising *limma* objects. However, they do support the following 
modes:

    - Regression on a continuous variable (e.g. age)
    - Categorical (only binary supported) (e.g. `hasPSEN` mutation)
    - Survival (not relevant to our analysis). 

- Because of this *PathwayPCA* should be able to give us some complementary 
information/results. 

### 4. AES-PCA

**Has AD**

- Here we test whether pathways are enriched on the binary variable `genotype` 
using AES-PCA

```{r}
samples_genotype <- samples %>%
  dplyr::select(Sample, genotype) %>%
  dplyr::rename(response = genotype)

omics_genotype <- CreateOmics(
  assayData_df = tidyDat,
  pathwayCollection_ls = hallmarkCol,
  response = samples_genotype,
  respType = "categ"
)

# Perform enrichment testing using AES-PCA
test_genotype <- AESPCA_pVals(
  object = omics_genotype,
  numReps = 0,
  numPCs = 2,
  parallel = TRUE,
  numCores = 2,
  adjustpValues = TRUE,
  adjustment = c("Holm", "BH")
)

test_genotype$pVals_df 

sigSets <- test_genotype$pVals_df %>%
  dplyr::filter(FWER_Holm < 0.05)

sigSets

# Perform enrichment testing using Supervised PCA
# test_AD2 <- SuperPCA_pVals(
#   object = omics_AD,
#   numPCs = 2,
#   parallel = TRUE,
#   numCores = 2,
#   adjustpValues = TRUE,
#   adjustment = c("SidakSS", "BY")
# )
# test_AD2$pVals_df
```

### 5. Plots

 - **Reactive oxygen species pathway**: Samples with AD tend to have lower 
 expression of genes in this pathway compared to samples without AD. 

```{r}
test_genotype$PCs_ls$HALLMARK_INTERFERON_ALPHA_RESPONSE %>% 
  set_rownames(attributes(test_genotype$PCs_ls)$sampleIDs) %>%
  rownames_to_column("Sample") %>%
  left_join(samples, by = "Sample") %>%
  ggplot(aes(x = V1, fill = genotype)) + 
  geom_histogram(bins = 8, colour = "white") +
  labs(x = "PC1 value for each sample",
       y = "Count", 
       fill = "Has AD?") +
  scale_fill_manual(values = c("#bbbbbb", "orangered")) +
  ggtitle("Interferon Alpha Response")
```

- All significant gene sets

```{r}
test_genotype$PCs_ls[sigSets$pathways] %>% 
  lapply(function(x){
    x %>% set_rownames(attributes(test_genotype$PCs_ls)$sampleIDs) %>%
      rownames_to_column("Sample")
  }) %>%
  bind_rows(.id = "pathway") %>%
  dplyr::mutate(pathway = gsub(x = pathway,
                               pattern = "_",
                               replacement = " "),
                pathway = gsub(x = pathway,
                               pattern = "HALLMARK ",
                               replacement = "")) %>%
  left_join(samples, by = "Sample") %>%
  ggplot(aes(x = V1, fill = genotype)) + 
  geom_histogram(bins = 8, colour = "white") +
  labs(x = "PC1 value for each sample",
       y = "Count", 
       fill = "Genotype") +
  scale_fill_manual(values = c( "orangered", "#bbbbbb")) +
  facet_wrap(~pathway, scales = "free_y")
  
```

