---
title: "Testing ProjectR on Zebrafish Dataset - Part 2"
author: "Nhi Hin"
date: "2020-05-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r include=FALSE}
library(CoGAPS)
library(projectR)
library(here)
library(tibble)
library(ggplot2)
library(dplyr)
library(genefilter)
library(edgeR)
library(pheatmap)
library(sva)
library(magrittr)
theme_set(theme_bw())
```


## Introduction

Due to the lack of results in our initial test of **CoGAPS** and 
**ProjectR**, the following modifications will be attempted:

- Using the *voom* matrix weight-adjusted counts. We noticed that 
two of the samples had much higher expression than other ones, and 
possibly that normalisation hadn't really corrected for that. 
The voom matrix has been adjusted with quality-weights, so that 
should help. 

- Filtering out genes with low variance as these are unlikely to 
contribute anyway. 

## 1. Import data

```{r}
dge_zeb <- readRDS(here("data", "datasets", "zebrafish", "dge.rds"))
dge_zeb

counts_nn <- dge_zeb %>% cpm(log=TRUE) 

# Are there values which are < 0?
sum(counts_nn < 0)

# majority of values are > 0
sum(counts_nn > 0)

# Filter out the rows containing negative values as these are likely 
# to be low expressed anyway, and the matrix factorisation used in 
# Cogaps will not accept them.
keepTheseGenes <- !rowSums(counts_nn < 0)
dge_zeb_f <- dge_zeb[keepTheseGenes,,keep.lib.sizes = FALSE] # 16,847 genes left

```

- Apply more aggressive filtering

```{r}
geneVars <- rowVars(cpm(dge_zeb_f, log=TRUE))

summary(geneVars)

# Filter out genes with variance below median variance 
keepTheseGenes2 <- rowVars(cpm(dge_zeb_f, log=TRUE)) > summary(geneVars)[[3]]

dge_zeb_f <- dge_zeb_f[keepTheseGenes2,,keep.lib.sizes = FALSE] 
```

- Import voom object

```{r}
voomData <- readRDS(here("data", "datasets", "zebrafish", "voomData.rds"))

voomExpression <- voomData$E%>%
  as.data.frame() %>%
  rownames_to_column("ensembl_gene_id") %>%
  dplyr::filter(ensembl_gene_id %in% rownames(dge_zeb_f$counts)) %>%
  column_to_rownames("ensembl_gene_id")

# get rid of negative values
keepTheseGenes <- !rowSums(voomExpression < 0)
voomExpression <- voomExpression[!rowSums(voomExpression < 0),]
```

## 2. Running CoGAPS

- Run CoGAPS with default settings 

```{r}
cogaps_def <- CoGAPS(as.matrix(voomExpression), nIterations = 1000)
```

- Quick heatmap to view the patterns (latent variables) found by CoGAPs. 
(Default number of patterns = 7)

```{r}
cogaps_def@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- Results here are already better and appear more interpretable than our 
initial test.

    - **Pattern 2** shows the outlier samples
    
    - **Pattern 7** captures aging (difference between 6 and 24 months)
    
    - **Pattern 1** captures difference between normoxia and hypoxia. 
    It seems that the 6 month mutants respond to hypoxia in a different 
    way to the 6 month WTs. 
    
    - **Pattern 4, 3, 5** aren't as easily interpretable, and there is some 
    noise. 
  

## 3. Varying the number of latent variables

- The default number of LVs is 7, but below we will try various numbers of LVs 
to see if it is possible to find one which accurately describes the entire 
experimental design including differences between mutant and wild types. 

```{r}
params <- new("CogapsParams")
params # defaultparameters

# Quick sweep to see how many patterns / LVs is ideal
# Each run takes 2-3 mins to run so below takes <20 mins
npatterns <- c(3,5,7,10,12,15,18,25) 
cogaps_nPatterns <- list()
for (i in npatterns){
  params <- setParam(params, "nPatterns", i) #change the number of patterns/LVs
  cogaps_nPatterns[i] <- CoGAPS(as.matrix(voomExpression),
                                params, nIterations = 1000)
}
```

- Plotting heatmaps starting with lowest `nPatterns = 3`

```{r}
cogaps_nPatterns[[3]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 5`

```{r}
cogaps_nPatterns[[5]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 10` 

```{r}
cogaps_nPatterns[[10]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 12`

```{r}
cogaps_nPatterns[[12]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 15`

```{r}
cogaps_nPatterns[[15]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 18`

```{r}
cogaps_nPatterns[[18]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

- `nPatterns = 25`

```{r}
cogaps_nPatterns[[25]]@sampleFactors %>% t %>% pheatmap(cluster_cols = FALSE)
```

