---
title: "zscoreCombined"
author: "Nhi Hin"
date: "2020-06-17"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
save_pheatmap_pdf <- function(x, filename, width=8, height=11) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}

palette <- colorRampPalette(c("#10c1e5", "#82e0b4",
                              "#F9F9F9", "#FBB829", "#FF0066"))(100)
```


## Introduction

- Aim here is to project all datasets together. 
- We will do an initial test with only the human datasets.

## 1. Import data

- Import the PC1 values from PathwayPCA for all datasets.

```{r}
# Antonell et al. dataset
samples_fAD <- readRDS(here("data", "datasets", "fad", "samples.rds"))
pc1_fAD <- readRDS(here("data", "datasets", 
                                "fad", "PC1_allSamples.rds"))

# Mayo Clinic RNA-seq (sporadic) AD dataset
samples_mayo <- readRDS(here("data", "datasets", "mayo", 
                                 "heatmapAnnot.rds"))
pc1_mayo <- readRDS(here("data", "datasets", "mayo", 
                               "PC1_TCsamples.rds"))
```

- The ordering of the samples (columns) in the PC1 tables needs to adjusted 
so that samples are ordered according to their condition/diagnosis. 

```{r}
samples_fAD %<>% dplyr::arrange(hasAD, group, apoe) %>%
  dplyr::mutate(Cluster = 6)
pc1_fAD %<>% magrittr::extract(samples_fAD$Sample,)

samples_mayo %<>% rownames_to_column("Sample") %>%
  dplyr::mutate(Diagnosis = gsub(x=Diagnosis,
                                 pattern = "Control",
                                 replacement = "control")) %>%
  dplyr::arrange(Cluster, Diagnosis, ApoE)
pc1_mayo %<>% magrittr::extract(samples_mayo$Sample,)
```


## 2. Scale Data

- All datasets will be standardised (mean removal and variance scaling) 
within each dataset. 

```{r}
# zscore <- function(x){
#   x <- (x-mean(x))/sd(x)
# }

# Dataset specific mean and standard dev.
mean(as.matrix(pc1_fAD))
sd(as.matrix(pc1_fAD))

scaleData <- function(x){
  x <- as.matrix(x)
  mean <- mean(x)
  sd <- sd(x)
  
  scaled_x <- (x-mean)/sd
  return(scaled_x)
}

scaled_pc1_fAD <- pc1_fAD %>%
  scaleData %>%
  set_rownames(rownames(pc1_fAD)) 

scaled_pc1_mayo <- pc1_mayo %>%
  scaleData %>%
  set_rownames(rownames(pc1_mayo))

# scaled_pc1_fAD <- pc1_fAD %>%
#   sapply(zscore) %>%
#   set_rownames(rownames(pc1_fAD))
# 
# scaled_pc1_mayo <- pc1_mayo %>%
#   sapply(zscore) %>%
#   set_rownames(rownames(pc1_mayo))



```


## 3. Initial heatmap

- To create a heatmap, we need the combined PC1 data as well as a 
combined annotation. 

```{r}
combinedAnnot <- samples_fAD %>%
  dplyr::select(Sample, Cluster, group, apoe, age) %>%
  dplyr::rename(Diagnosis = group,
                ApoE = apoe,
                AgeAtDeath = age) %>%
  dplyr::mutate(apoe = gsub(x =ApoE,
                            pattern = "_",
                            replacement = "")) %>%
  dplyr::select(Sample, Cluster, Diagnosis, AgeAtDeath, ApoE) %>%
  bind_rows(samples_mayo) %>%
  dplyr::mutate(Diagnosis = as.factor(Diagnosis)) %>%
  column_to_rownames("Sample")

combinedPC1 <- rbind(scaled_pc1_fAD, scaled_pc1_mayo) %>%
  magrittr::extract(rownames(combinedAnnot),)
colnames(combinedPC1) %<>% 
  gsub(x = ., pattern = "HALLMARK_", replacement = "") %>%
  gsub(x = ., pattern = "_", replacement = " ")

#pheatmap(pc1_fAD, color = palette, scale="none")

combinedHeatmap <- pheatmap(combinedPC1%>%t,
         color = palette,
         cluster_cols = FALSE,
         annotation_col = combinedAnnot,
         scale="none"
)

# save_pheatmap_pdf(combinedHeatmap, here("data", "PC1", "fig",
#                                         "fADsADheatmap.pdf"),
#                   width = 15, height = 12)
```

## 4. Zebrafish and Mouse Data

- Import in PC1 values

```{r}
# Zebrafish data
pc1_zeb <- readRDS(here("data", "datasets", "zebrafish", 
                                "PC1.rds"))

# Mouse dataset 1
pc1_mouse1<- readRDS(here("data", "datasets", 
                          "mouse", "GSE140286", "PC1.rds"))
rownames(pc1_mouse1) <- gsub(x = rownames(pc1_mouse1),
                             pattern = "^(.*)$",
                             replacement = "\\1_D1")

# Mouse dataset 2
pc1_mouse2 <- readRDS(here("data", "datasets", "mouse", 
                                "GSE142633", "PC1.rds"))
rownames(pc1_mouse2) <- gsub(x = rownames(pc1_mouse2),
                             pattern = "^(.*)$",
                             replacement = "\\1_D2")
```

- Sample Information can be extracted from previously saved DGEList objects.

```{r}
samples_zeb <- readRDS(here("data", "datasets", "zebrafish", 
                        "dge.rds"))$samples

samples_mouse1 <- readRDS(here("data", "datasets", "mouse", 
                        "GSE140286", "dge.rds"))$samples

samples_mouse2 <- readRDS(here("data", "datasets", "mouse", 
                        "GSE142633", "dge.rds"))$samples
```

- The ordering of samples needs to be adjusted so they are ordered 
according to their condition. 

```{r}
samples_zeb %<>% rownames_to_column("Sample") %>%
  dplyr::arrange(Age, Hypoxia, Genotype)%>%
  dplyr::filter(Hypoxia==0)
  
pc1_zeb %<>% magrittr::extract(samples_zeb$Sample,) 

samples_mouse1 %<>% rownames_to_column("Sample") %>%
  dplyr::arrange(age, genotype)%>%
  dplyr::mutate(Sample = paste0(Sample, "_D1"))
pc1_mouse1 %<>% magrittr::extract(samples_mouse1$Sample,)

samples_mouse2 %<>% rownames_to_column("Sample") %>%
  dplyr::arrange(age, genotype) %>%
  dplyr::mutate(Sample = paste0(Sample, "_D2"))
pc1_mouse2 %<>% magrittr::extract(samples_mouse2$Sample,)
```

- Scale data

```{r}
scaled_pc1_zeb <- pc1_zeb[1:8,] %>%
  scaleData %>%
  rbind(scaleData(pc1_zeb[9:16,])) %>%
  set_rownames(rownames(pc1_zeb))

scaled_pc1_mouse1 <- pc1_mouse1 %>%
  scaleData %>%
  set_rownames(rownames(pc1_mouse1))

scaled_pc1_mouse2 <- pc1_mouse2 %>%
  scaleData %>%
  set_rownames(rownames(pc1_mouse2))
```

- Combined heatmap 

```{r}
combinedAnnot2 <- samples_zeb %>%
  dplyr::select(Sample, Age, Genotype) %>%
  dplyr::rename(age = Age, 
                genotype = Genotype) %>%
  dplyr::mutate(age = as.numeric(as.character(age)),
                genotype=gsub(x=genotype, 
                              pattern="wt",
                              replacement = "WT")) %>%
  bind_rows(samples_mouse1[, c("Sample", "age", "genotype")],
            samples_mouse2[, c("Sample","age", "genotype")]%>%
              dplyr::mutate(age=as.numeric(as.character(age))),
              .id = "Dataset") %>%
  dplyr::mutate(Dataset = as.numeric(as.character(Dataset)),
                age = as.numeric(as.character(age)),
                genotype = as.character(genotype))%>%
    column_to_rownames("Sample")
  

combinedPC12 <- rbind(scaled_pc1_zeb, scaled_pc1_mouse1, scaled_pc1_mouse2)

combinedHeatmap3 <- pheatmap(combinedPC12%>%t,
         color = palette,
         cluster_cols = FALSE,
         gaps_col = c(16, 16, 16, 31,31,31,
                      8, 20,26,37),
         annotation_col = combinedAnnot2,
         scale="none"
)

save_pheatmap_pdf(combinedHeatmap3, here("data", "PC1", "fig",
                                        "zebrafishMouseHeatmap.pdf"),
                  width = 15, height = 12)
```


