---
title: "zscoreCombined2"
author: "Nhi Hin"
date: "2020-06-29"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r functions}
# Scale data by global / dataset specific mean and standard deviation. 
scaleData <- function(x){
  x <- as.matrix(x)
  mean <- mean(x)
  sd <- sd(x)
  
  scaled_x <- (x-mean)/sd
  return(scaled_x)
}

# Save pheatmap as a pdf in A4 default size. 
save_pheatmap_pdf <- function(x, filename, width=8, height=11) {
   stopifnot(!missing(x))
   stopifnot(!missing(filename))
   pdf(filename, width=width, height=height)
   grid::grid.newpage()
   grid::grid.draw(x$gtable)
   dev.off()
}
```


## 1. Introduction

- Aim here is to combine the Mouse and Zebrafish datasets. 

- Import in the PC1 values

```{r}
pc1_zeb <- list(
  age_6mths = readRDS(here("data", "datasets", "zebrafish",
                           "PC1_6mths.rds")),
  age_24mths = readRDS(here("data", "datasets", "zebrafish",
                            "PC1_24mths.rds"))
)

pc1_mouse1 <- list(
  age_3mths = readRDS(here("data", "datasets", "mouse", "GSE140286",
                           "PC1_3mths.rds")),
  age_6mths = readRDS(here("data", "datasets", "mouse", "GSE140286",
                           "PC1_6mths.rds")),
  age_12mths = readRDS(here("data", "datasets", "mouse", "GSE140286",
                           "PC1_12mths.rds"))
)

pc1_mouse2 <- list(
  age_6mths = readRDS(here("data", "datasets","mouse",
                           "GSE142633","PC1_6mths.rds")),
  age_11mths = readRDS(here("data", "datasets","mouse",
                           "GSE142633","PC1_11mths.rds"))
)
```

- Load in the sample information from previously created DGEList objects.

```{r}
# Zebrafish dataset
samples_zeb <- readRDS(here("data", "datasets", "zebrafish", 
                        "dge.rds"))$samples
samples_zeb$Genotype %<>% as.factor %>% relevel("wt")
samples_zeb <- list(
  age_6mths = samples_zeb %>% dplyr::filter(Age == 6,
                                            Hypoxia == 0),
  age_24mths = samples_zeb %>% dplyr::filter(Age == 24,
                                             Hypoxia == 0)
)

# Mouse dataset
samples_mouse1 <- readRDS(here("data", "datasets", "mouse", 
                        "GSE140286", "dge.rds"))$samples %>%
  rownames_to_column("Sample")
samples_mouse1$genotype %<>% relevel("WT")
samples_mouse1 <- list(
  age_3mths = samples_mouse1 %>% dplyr::filter(age == 3),
  age_6mths = samples_mouse1 %>% dplyr::filter(age == 6),
  age_12mths = samples_mouse1 %>% dplyr::filter(age == 12)
)

samples_mouse2 <- readRDS(here("data", "datasets", "mouse", 
                        "GSE142633", "dge.rds"))$samples %>%
  rownames_to_column("Sample")
samples_mouse2$genotype %<>% as.factor %>% relevel("WT")
samples_mouse2 <- list(
  age_6mths = samples_mouse2 %>% dplyr::filter(age == 6),
  age_11mths = samples_mouse2 %>% dplyr::filter(age == 11)
)
```

- The ordering of samples needs to be adjusted so they are ordered 
according to their condition. 

```{r}
# Zebrafish samples are already in order but we will just select the 
# relevant columns
samples_zeb %<>%
  lapply(function(x){
    x %>%
      dplyr::rename(Sample = sample,
                    age = Age,
                    genotype = Genotype) %>%
      dplyr::select(Sample, age, genotype)
  })
pc1_zeb$age_6mths %<>% magrittr::extract(samples_zeb$age_6mths$Sample,) 
pc1_zeb$age_24mths %<>% magrittr::extract(samples_zeb$age_24mths$Sample,) 

# Mouse samples
samples_mouse1 %<>%
  lapply(function(x){
    x %<>% dplyr::mutate(genotype = gsub(x = genotype,
                                        pattern = "WT",
                                        replacement = "wt"))
    x$genotype %<>% as.factor %>% relevel("wt") 
    x %<>% dplyr::arrange(genotype) %>%
      dplyr::select(Sample, age, genotype) %>%
      dplyr::mutate(age = as.factor(age))
  })
pc1_mouse1$age_3mths %<>% magrittr::extract(samples_mouse1$age_3mths$Sample,) 
pc1_mouse1$age_6mths %<>% magrittr::extract(samples_mouse1$age_6mths$Sample,) 
pc1_mouse1$age_12mths %<>% magrittr::extract(samples_mouse1$age_12mths$Sample,) 

samples_mouse2 %<>%
  lapply(function(x){
    x %<>% dplyr::mutate(genotype = gsub(x = genotype,
                                        pattern = "WT",
                                        replacement = "wt"))
    x$genotype %<>% as.factor %>% relevel("wt") 
    x %<>% dplyr::arrange(genotype) %>%
      dplyr::select(Sample, age, genotype) %>%
      dplyr::mutate(age = as.factor(age))
  })
pc1_mouse2$age_6mths %<>% magrittr::extract(samples_mouse2$age_6mths$Sample,) 
pc1_mouse2$age_11mths %<>% magrittr::extract(samples_mouse2$age_11mths$Sample,) 
```

## 2. Scale Data

- All datasets will be standardised (mean removal and variance scaling) 
within each dataset. 

```{r}
scaled_pc1_zeb <- pc1_zeb %>% lapply(scaleData) 
 
scaled_pc1_mouse1 <- pc1_mouse1 %>% lapply(scaleData) 

scaled_pc1_mouse2 <- pc1_mouse2 %>% lapply(scaleData)
```

## 3. Heatmaps

- To create a combined heatmap, we need to make a combined PC1 matrix, 
combined annotation data.frame. 

- As the lifespan of mice can be considered roughly half of zebrafish, 
we will firstly plot the 24 month old zebrafish and 12 month old mice on 
the same heatmap to see how comparable they are. 

```{r, warning=FALSE}
palette <- colorRampPalette(c("#10c1e5", "#82e0b4",
                              "#F9F9F9", "#FBB829", "#FF0066"))(100)

combinedAnnot_zebMouse <- samples_zeb$age_24mths %>%
  bind_rows(samples_mouse1$age_12mths)%>%
  bind_rows(samples_mouse2$age_11mths) %>%
  as_tibble %>%
  mutate(age = as.factor(as.numeric(age)),
         genotype = as.factor(genotype))%>%
    column_to_rownames("Sample")

combinedPC1_zebMouse <- rbind(scaled_pc1_zeb$age_24mths, 
                     scaled_pc1_mouse1$age_12mths,
                     scaled_pc1_mouse2$age_11mths) 

colnames(combinedPC1_zebMouse) %<>%
  gsub(x = ., pattern = "HALLMARK_", replacement = "")
colnames(combinedPC1_zebMouse) %<>%
  gsub(x = ., pattern = "_", replacement = " ")

combinedHeatmap_zebMouse <- pheatmap(combinedPC1_zebMouse%>%t,
         color = palette,
         cluster_cols = FALSE,
         cellwidth = 10,
         cellheight = 10,
         gaps_col = c(8,13),
         border_color = "white",
         annotation_col = combinedAnnot_zebMouse,
         scale="none"
)

```

## 4. t-tests for significant differences

- Test if there are significant differences between mutant and wt samples 
in each of the datasets.

```{r}
# Input
#   x: A matrix/data.frame of PC1 values where rownames are samples 
#      and colnames are genesets. 
#   group: A factor vector of the group which each sample (row) in x 
#          belongs to (e.g. genotype). Ordering must be the same as the 
#          order of rownames in x.

testPC1DifferentBetween2Groups <- function(x, y, padj = "fdr"){
  # format x to be in long form and add the group information
  pc1_data <- x %>% 
    as.data.frame %>%
    tibble::rownames_to_column("Sample") %>%
    dplyr::mutate(group = y)%>%
    melt(id.vars = c("Sample", "group"))
  
  # perform a t-test for each gene set between the 2 groups
  t_pc1_data <- lm(value ~ group + variable, data = pc1_data)
  
  # put results into a list, including the model as 
  # well as the raw and adjusted p-values
  results <- list(
    t = t_pc1_data,
    p = summary(t_pc1_data)$coefficients[,4] %>%
      as.data.frame %>%
      rownames_to_column("Geneset") %>%
      set_colnames(c("Geneset", "p")) %>%
      dplyr::mutate(adj_p = p.adjust(p, "fdr"),
                    Geneset = gsub(x = Geneset,
                                   pattern = "variable",
                                   replacement = ""))%>%
      dplyr::arrange(p) %>%
      dplyr::mutate(sig = adj_p < 0.05)
  )
  
  return(results)
}



test2 <- testPC1DifferentBetween2Groups(pc1_zeb$age_6mths, 
                                        samples_zeb$age_6mths$genotype)

test3 <- testPC1DifferentBetween2Groups(pc1_zeb$age_24mths, 
                                        samples_zeb$age_24mths$genotype)

test4 <- testPC1DifferentBetween2Groups(pc1_mouse2$age_11mths,
                                        samples_mouse2$age_11mths$genotype)

# data_zeb_6mths <- pc1_zeb$age_6mths %>% as.data.frame %>%
#   rownames_to_column("Sample") %>%
#   melt %>% 
#   left_join(samples_zeb$age_6mths[, c("Sample", "genotype")]) 
# 
# t_zeb_6mths <- lmer(value ~ genotype + variable + (genotype|variable), data=data_zeb_6mths) 
# t_zeb_6mths_p <- summary(t_zeb_6mths)$coefficients[,4] %>% 
#   as.data.frame()%>%
#   rownames_to_column("Geneset") %>%
#   set_colnames(c("Geneset", "p"))%>%
#   dplyr::mutate(adj_p = p.adjust(p, "fdr"))%>%
#   dplyr::arrange(p) %>%
#   dplyr::mutate(sig = adj_p < 0.05)
```


```{r}
testing <- combinedPC1_zebMouse[, 5] %>% 
  enframe(name = "sample") %>%
  mutate(dataset = case_when(str_detect(sample, "ale") ~ "mouse", 
                             !str_detect(sample, "ale") ~ "zebrafish"))%>%
  mutate(genotype = case_when(str_detect(sample, "[Ww]") ~ "WT", 
                              !str_detect(sample, "[Ww]") ~ "Mutant")) 
testing$genotype %<>% as.factor %>% relevel("WT")
  
  testing %>% with(lmer(value ~ genotype + (1|dataset))) %>%
    summary() %>% coef()
  %>% as.data.frame() %>% 
    rownames_to_column("coef") %>% as_tibble()
```

```{r}
zebAndMouse <- combinedPC1_zebMouse %>% as.data.frame %>%
  rownames_to_column("Sample") %>%
  left_join(combinedAnnot_zebMouse %>%
              rownames_to_column("Sample") %>%
              mutate(dataset = c(rep("zebrafish", 8),
                                 rep("mouse1", 5),
                                 rep("mouse2", 6)))) %>%
  dplyr::mutate(group = case_when(
    genotype == "q96" ~ "fAD",
    genotype == "5XFAD" ~ "fAD",
    genotype == "wt" ~ "wt"
  )) %>%
  dplyr::mutate(group = as.factor(group)) %>%
  dplyr::select(-genotype, -age)
zebAndMouse$group %<>% relevel("wt")

zebAndMouse <- zebAndMouse %>%
  melt(id.vars = c("Sample", "group", "dataset")) # %>%
#with(lmer(value ~ group + (1|dataset))) %>%
#  summary()

zebAndMouse_split <- zebAndMouse %>% split(.$variable)
resTest <- zebAndMouse_split %>% lapply(function(x){
  x %>% with(lmer(value ~ group + (1|dataset))) %>%
    summary %>%
    coef()%>%
    as.data.frame %>%
    rownames_to_column("coef") %>%
    as_tibble
})
  
```

```{r}
scaled_pc1_zeb$age_24mths %>%
  as.data.frame %>%
  rownames_to_column("Sample") %>%
  left_join(samples_zeb$age_24mths, by = "Sample")%>%
  dplyr::select(-age)%>%
  melt(id.vars = c("Sample", "genotype")) %>%
  with(lmer(value ~ genotype + (1|variable)))%>%
  summary()
```





- As the oldest age groups of the zebrafish and mice available, it would 
also be worth plotting them alongside the human data. 

- We will import the combined annotation and PC1 values from the 
fAD and Mayo datasets below. 

```{r}
combinedAnnot_human <- readRDS(here("data", "PC1", "combinedAnnot.rds")) %>%
  rownames_to_column("Sample") %>%
  dplyr::select(Sample, Cluster, Diagnosis) %>%
  dplyr::rename(Group = Diagnosis)

combinedPC1_human <- readRDS(here("data", "PC1", "combinedPC1.rds"))


```

- Full combined

```{r}
combinedAnnot_zebMouse2 <- combinedAnnot_zebMouse %>%
  rownames_to_column("Sample") %>%
  dplyr::mutate(Cluster = c(rep(7,8), rep(8,5)))%>%
  dplyr::rename(Group = genotype) %>%
  dplyr::select(Sample, Cluster, Group)

combinedAnnot_all <- bind_rows(combinedAnnot_zebMouse2, 
                              combinedAnnot_human)%>%
  dplyr::mutate(Group = as.character(Group),
                Cluster = as.numeric(Cluster))%>%
    column_to_rownames("Sample") 

combinedPC1_all <- rbind(combinedPC1_zebMouse,
                         combinedPC1_human)

combinedHeatmap_all <- pheatmap(
  combinedPC1_all %>% t,
  cluster_cols = FALSE,
  gaps_col = c(8,8+5,13+21),
  color = palette,
  annotation_col = combinedAnnot_all
)

combinedHeatmap_all %>% save_pheatmap_pdf(
  here::here("data", "PC1", "fig", "combinedHeatmap_all.pdf"), width=11
)
```

